---
title: "Benefits of using LF-LAM in the diagnosis of TB"
format: html
editor: visual
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
echo = FALSE,
message = FALSE,
warning = FALSE,
cache = FALSE
)

# Load required packages
pacman::p_load(
  rio,
  here,
  tidyverse,
  janitor,
  geepack,
  aod,
  gtsummary,
  labelled
)

theme_gtsummary_journal(journal = "jama")
theme_gtsummary_compact()
```

```{r data}
dfall <- 
  
  # Import & merge all excel files
  import(here('data', 'table1.xlsx')) %>%
  select(-c(Remark:updated_at)) %>%
  mutate(id = as.numeric(id)) %>%
  left_join(import(here('data', 'table2.xlsx')) %>%
              select(-c(id, Remark:updated_at), id = patient_id),
            by = 'id') %>%
  left_join(import(here('data', 'table3.xlsx')) %>%
              select(-c(id, Remark:updated_at), id = patient_id), 
            by = 'id') %>%
  left_join(import(here('data', 'table4.xlsx')) %>%
              select(-c(id, Remark:updated_at), id = patient_id), 
            by = 'id') %>%
  left_join(import(here('data', 'table5.xlsx')) %>%
              select(-c(id, Remark:updated_at), id = patient_id), 
            by = 'id') %>%
  left_join(import(here('data', 'table6.xlsx')) %>%
              select(-c(id, Remark:updated_at), id = patient_id), 
            by = 'id') %>%
  left_join(import(here('data', 'table7.xlsx')) %>%
              select(-c(id, created_at:isDeleted), id = patient_id) %>%
              group_by(id) %>%
              mutate(visit = row_number()) %>%
              ungroup() %>%
              pivot_wider(
                names_from = visit,
                values_from = VirologyCheck:ViralLoad,
                names_vary = 'slowest'), 
            by = 'id') %>%
  left_join(import(here('data', 'table8.xlsx')) %>%
              select(-c(id, Remark:updated_at), id = patient_id), 
            by = 'id') %>%
  left_join(import(here('data', 'table843.xlsx')) %>%
              select(-c(id, Remark:updated_at), id = patient_id), 
            by = 'id') %>%
  left_join(import(here('data', 'table9.xlsx')) %>%
              select(-c(id, Remark:updated_at), id = patient_id), 
            by = 'id') %>%
  left_join(import(here('data', 'table10.xlsx')) %>%
              select(-c(id, Remark:updated_at), id = patient_id), 
            by = 'id') %>%
  left_join(import(here('data', 'table11.xlsx')) %>%
              select(-c(id, Remark:updated_at), id = patient_id), 
            by = 'id') %>%
  
  # Simplify variable names
  clean_names() %>%
  
  # Summary of multiple variables (5 tb symptom 5, 5 lab results, 14 antibiotics)
  rowwise() %>%
  mutate(
    # TB symptoms
    tb_symptom = max(
      c_across(num_range("criteria_2_1_", 1:5)) == 1, 
      na.rm = TRUE
    ),
    # AFB
    afb = max(
      c_across(num_range("afb_check_result", 1:5)) == 2, 
      na.rm = TRUE
    ),
    # Culture
    culture =max(
      c_across(num_range("tb_culture_check_result", 1:5)) == 2,
      na.rm = TRUE
    ),
    # Xpert
    xpert = max(
      c_across(num_range("xpert_check_mtb_result", 1:5)) == 1,
      na.rm = TRUE
    ),
    # Number of antibiotics
    atb_num = sum(
      c_across(c(empirical_antibiotic_drug_amoxicillin:empirical_antibiotic_drug_other) & 
                 starts_with('empirical_antibiotic_drug_')),
      na.rm = TRUE
    ),
    # Antibiotic start date
    atb_start = min(
      c_across(c(amoxicillin_start_date:other_start_date) & ends_with('_start_date')),
      na.rm = TRUE
    ),
    # Antibiotic stop date
    atb_stop = max(
      c_across(c(amoxicillin_end_date:other_end_date) & ends_with('_end_date')),
      na.rm = TRUE
    ),
  ) %>%
  ungroup() %>%
  
  # Key variables
  mutate(
    tb_symptom = na_if(tb_symptom, -Inf),
    afb = na_if(afb, -Inf),
    culture = na_if(culture, -Inf),
    xpert = na_if(xpert, -Inf),
    # TB definition
    tb_def = case_when(
      # Definite TB
      culture == 1 | xpert == 1                                                    ~ 1,
      # Probable TB
      (tb_symptom == 1 | afb == 1 | cxr_check_result == 2 | ct_check_result == 2) & 
       antituberculosis_drugs_receive == 1 & tb_diagg_result != 5                  ~ 2,
      # Not TB
      TRUE                                                                         ~ 0 
    ),
    # TB diagnosis (mutually exclusive)
    tb_diag = case_when(
      disseminated_tb == 1                        ~ 3,
      extra_pulmonary_tb == 1 & pulmonary_tb == 1 ~ 3,
      extra_pulmonary_tb == 1                     ~ 2,
      pulmonary_tb == 1                           ~ 1,
      non_tb == 1                                 ~ 0,
      TRUE                                        ~ NA_real_
    ),
    # TB yes/no
    tb = factor(
      case_when(
        tb_def %in% 1:2 ~ 1,
        TRUE            ~ 0
      ),
      levels = c(1, 0),
      labels = c('Yes', 'No')
    ),
    # LAM +/-
    lam = factor(
      active_tb_alere_determine_test_result,
      levels = c(2, 1),
      labels = c('Positive', 'Negative')
    ),
    
    # Baseline characteristics
    across(where(is.POSIXct), ~ as.Date(.)),
    ageatvisit = as.numeric(difftime(service_date, dob, units = "days")) / 365.25,
    sex = factor(
      sex,
      levels = c(1, 2),
      labels = c('Male', 'Female')
    ),
    nationgrp = factor(
      case_when(nation == 1                ~ 1,
                nation %in% c(2, 3, 4, 99) ~ 2,
                TRUE                       ~ NA_real_),
      levels = c(1, 2),
      labels = c('Thai', 'Non-Thai')
    ),
    lengthart = as.numeric(difftime(
      service_date,
      coalesce(arv_start_date, arv_start_date2),
      units = "days")) / 365.25,
    
    # Factors
    atb = factor(
      empirical_antibiotic_drug_receive,
      levels = 1:0,
      labels = c('Yes', 'No')
    ),
    atb_time = na_if(as.numeric(
      difftime(atb_stop, atb_start, units = "days")
    ),-Inf),
    atb_timegrp = cut(
      atb_time,
      breaks = c(0, 4, Inf),
      labels = c('< 5', '>= 5'),
      include.lowest = TRUE
    ),
    atb_numgrp = cut(
      atb_num,
      breaks = c(0, 1, 2, Inf),
      labels = c('1', '2', '>= 3')
    ),
    tx_time = as.numeric(
      difftime(
        antituberculosis_drugs_start_date,
        active_tb_alere_determine_test_date,
        units = "days"
      )
    ),
    tx_timegrp = cut(
      tx_time,
      breaks = c(-Inf,-1, 3, 22, Inf),
      labels = c(
        'Start TB treatment before LAM result',
        '0-3 days',
        '4-22 days',
        '>22 days'
      )
    ),
    h_time = as.numeric(
      difftime(
        discharged_date,
        service_date,
        units = "days"
      )
    ),
    h_timegrp = cut(
      h_time,
      breaks = c(-1, 3, 6, Inf),
      labels = c(
        '0-3 days',
        '4-6 days',
        '>=7 days'
      )
    ),
    spu_cul = case_when(
      (sample_type1 == 1 & tb_culture_check_result1 == 2) |
      (sample_type2 == 1 & tb_culture_check_result2 == 2) |
      (sample_type3 == 1 & tb_culture_check_result3 == 2) |
      (sample_type4 == 1 & tb_culture_check_result4 == 2) |
      (sample_type5 == 1 & tb_culture_check_result5 == 2) ~ 1,
      (sample_type1 == 1 & tb_culture_check_result1 %in% c(1,3)) |
      (sample_type2 == 1 & tb_culture_check_result2 %in% c(1,3)) |
      (sample_type3 == 1 & tb_culture_check_result3 %in% c(1,3)) |
      (sample_type4 == 1 & tb_culture_check_result4 %in% c(1,3)) |
      (sample_type5 == 1 & tb_culture_check_result5 %in% c(1,3)) ~ 0,
      TRUE ~ NA_real_
    ),
    bld_cul = case_when(
      (sample_type1 == 3 & tb_culture_check_result1 == 2) |
      (sample_type2 == 3 & tb_culture_check_result2 == 2) |
      (sample_type3 == 3 & tb_culture_check_result3 == 2) |
      (sample_type4 == 3 & tb_culture_check_result4 == 2) |
      (sample_type5 == 3 & tb_culture_check_result5 == 2) ~ 1,
      (sample_type1 == 3 & tb_culture_check_result1 %in% c(1,3)) |
      (sample_type2 == 3 & tb_culture_check_result2 %in% c(1,3)) |
      (sample_type3 == 3 & tb_culture_check_result3 %in% c(1,3)) |
      (sample_type4 == 3 & tb_culture_check_result4 %in% c(1,3)) |
      (sample_type5 == 3 & tb_culture_check_result5 %in% c(1,3)) ~ 0,
      TRUE ~ NA_real_
    ),
    txres = factor(treatment_result,
                  levels = c(1:4, 99),
                  labels = c('Improve', 'Not improve', 'Refer', 'Dead', 'Other')),
    dead = factor(ifelse(treatment_result == 4, 1, 0),
                  levels = c(1, 0),
                  labels = c('Dead', 'Alive'))
  ) %>%
  
  # Variable labels
  set_variable_labels(
    lam          = 'LF-LAM',
    ageatvisit   = "Age",
    sex          = "Sex",
    nationgrp    = "Nationality",
    absolute_cd4 = "CD4 (cell/μL)",
    lengthart    = 'Length in ART (year)',
    atb          = "Antibiotic given",
    atb_time     = "Time on antibiotic",
    atb_timegrp  = "Time on antibiotic",
    atb_numgrp   = "Number of antibiotic",
    tx_time      = 'Time of start of TB treatment after LAM test',
    tx_timegrp   = 'Time of start of TB treatment after LAM test',
    h_time       = 'Length of stay',
    h_timegrp    = 'Length of stay',
    txres        = 'Treatment result',
    dead         = 'Dead'
  )

dfinc <- dfall %>% 
  # Exclude incomplete routine TB diagnosis
  filter(!is.na(pulmonary_tb) | 
         !is.na(extra_pulmonary_tb) | 
         !is.na(disseminated_tb) |
         !is.na(non_tb)) %>%
  # Exclude conflicts between definition & diagnosis
  filter((tb_def == 0 & non_tb == 1) |
         (tb_def %in% 1:2 &
           (pulmonary_tb == 1 | extra_pulmonary_tb == 1 | disseminated_tb == 1))) %>%
  # Exclude missing lam 
  filter(!is.na(lam)) %>%
  # Exclude non-standard LF-LAM testing
  filter(active_tb_alere_determine_test_result %in% 1:2) %>% 
  # Exclude incomplete treatment information
  filter((!is.na(antituberculosis_drugs_receive) | !tb_diag == 1))

dfipd <- dfinc %>% 
  filter(patient_type == 1)

dfipdsnpt <- dfinc %>%
  filter(patient_type == 1 &
           afb == 0 &
           tb_diag %in% c(1,3))
```

**Flow diagram**

```{r flowchart}
source(here('code','flowchart.R'))
d
```

**Baseline characteristics of all IPD cases** (n=`r nrow(dfipd)`)

```{r ipd}
t1a <- tbl_summary(
  dfipd %>% select(ageatvisit, sex, nationgrp, absolute_cd4, lengthart),
  type = list(lengthart ~ 'continuous2'),
  statistic = list(lengthart ~ c(
    "{median} ({p25}, {p75})", "{mean} ({min}, {max})"
  ))
) %>%
  bold_labels()
t1a
```

**Baseline characteristics of IPD with sputum negative pulmonary TB** (n=`r nrow(dfipdsnpt)`)

```{r ipdsnpt}
t1b <- tbl_summary(
  dfipdsnpt %>% select(ageatvisit, sex, nationgrp, absolute_cd4, lengthart),
  type = list(lengthart ~ 'continuous2'),
  statistic = list(lengthart ~ c(
    "{median} ({p25}, {p75})", "{mean} ({min}, {max})"
  ))
) %>%
  bold_labels()
t1b
```

**Treatment and length of stay among IPD SNPT** (n=`r nrow(dfipdsnpt)`)

```{r table1}
t2a <- tbl_summary(
  dfipdsnpt %>%
    select(
      lam,
      atb,
      atb_time,
      atb_timegrp,
      atb_numgrp,
      tx_time,
      tx_timegrp,
      h_time,
      h_timegrp,
      txres,
      dead
    ),
  by = lam,
  type = list(all_continuous() ~ 'continuous2'),
  statistic = list(
    all_continuous() ~ c("{median} ({p25}, {p75})", "{mean} ({min}, {max})")
  )
) %>%
  bold_labels() %>%
  add_overall() %>%
  add_p() %>%
  bold_p()
t2a
```

**Treatment and length of stay among IPD SNPT who start TB treatment after LAM result** (n=`r nrow(filter(dfipdsnpt, tx_time >= 0))`)

```{r table2}
t2b <- tbl_summary(
  dfipdsnpt %>%
    filter(tx_time >= 0) %>% 
    select(
      lam,
      atb,
      atb_time,
      atb_timegrp,
      atb_numgrp,
      tx_time,
      tx_timegrp,
      h_time,
      h_timegrp,
      txres,
      dead
    ),
  by = lam,
  type = list(all_continuous() ~ 'continuous2'),
  statistic = list(
    all_continuous() ~ c("{median} ({p25}, {p75})", "{mean} ({min}, {max})")
  )
) %>%
  bold_labels() %>%
  add_overall() %>%
  add_p() %>%
  bold_p()
t2b
```

```{r result}
n1 = ipd
n2 = round(median(dfipd$ageatvisit, na.rm = TRUE), digits = 0)
n3 = nrow(filter(dfipd,sex=='Male'))
n4 = percent(n3 / n1, accuracy = 0.1)
ncd4 = nrow(filter(dfipd, absolute_cd4 < 200))
n5 = median(dfipd$absolute_cd4, na.rm = TRUE)
n6 = round(quantile(dfipd$absolute_cd4, probs = c(0,0.25,0.5,0.75,1), na.rm=TRUE)[2], digits = 0)
n7 = round(quantile(dfipd$absolute_cd4, probs = c(0,0.25,0.5,0.75,1), na.rm=TRUE)[4], digits = 0)
n8 = ipdsn
n9 = percent(n8 / n1, accuracy = 0.1)
n10 = ipdsp
n11 = percent(n10 / n1, accuracy = 0.1)
n12 = ipdsm
n13 = percent(n12 / n1, accuracy = 0.1)
n14 = nrow(filter(dfipd, afb == 0 & spu_cul == 1))
n15 = nrow(filter(dfipd, afb == 0 & !is.na(spu_cul)))
n16 = percent(n14 / n15, accuracy = 0.1)
n17 = nrow(filter(dfipd, afb == 0 & bld_cul == 1))
n18 = nrow(filter(dfipd, afb == 0 & !is.na(bld_cul)))
n19 = percent(n17 / n18, accuracy = 0.1)
n20 = nrow(filter(dfipd, afb == 0 & xpert == 1))
n21 = nrow(filter(dfipd, afb == 0 & !is.na(xpert)))
n22 = percent(n20 / n21, accuracy = 0.1)
n23 = nrow(filter(dfipd, afb == 0 & lam == 'Positive'))
n24 = nrow(filter(dfipd, afb == 0 & !is.na(lam)))
n25 = percent(n23 / n24, accuracy = 0.1)
n26 = nrow(filter(dfipd, afb == 0 & tb == 'Yes'))
n27 = percent(n26 / n8, accuracy = 0.1)
n28 = nrow(filter(dfipd, afb == 0 & tb == 'Yes' & tb_def == 1))
n29 = percent(n28 / n26, accuracy = 0.1)
n30 = nrow(filter(dfipd, afb == 0 & tb == 'Yes' & tb_def == 2))
n31 = percent(n30 / n26, accuracy = 0.1)
n32 = percent(ipdsnptpos / ipdsnpt, accuracy = 0.1)
n33 = ipdsnptpos
n34 = ipdsnpt
n35 = nrow(dfipdsnpt %>% filter(tx_time >= 0))
```

**Results**

Of `r n1` PLHIV admitted in IPD, `r n3` (`r n4`) were male, and the median age was `r n2`, and ``. Median CD4-count was `r n5` cells/mm3 \[interquartile range (IQR), `r n6`-`r n7` cells/mm3\], `r n8` (`r n9`) were sputum AFB-negative, `r n10` (`r n11`) were sputum AFB positive, and `r n12` missing (`r n13`). Of `r n8` PLHIV with sputum AFB negative, `r n14`/`r n15` (`r n16`) were sputum culture-positive, `r n17`/`r n18` (`r n19`) blood-culture positive, `r n20`/`r n21` (`r n22`) GeneXpert positive, and `r n23`/`r n24` (`r n25`) LAM-positive. Of `r n8`, `r n26` (`r n27`) were TB cases, including `r n28` (`r n29`) definite TB and `r n30` (`r n31`) probable TB. The LAM sensitivity was `r n32` (`r n33`/`r n34`) among IPD SNPT PLHIV. Of `r n26` cases, `r n34` were pulmonary or disseminated TB. We included `r n35` whose TB treatment was initiated after performing LF-LAM. Outcomes among IPD PLHIV with LF-LAM positive and negative are shown in the table.

```{r}
dfipd %>%
  mutate(
    on = factor(
      on_arv,
      levels = 0:3,
      labels = c('No', 'No, experienced and stopped ART', 'Yes', 'No, restarted ART')
    )
  ) %>%
  tbl_summary(
    include = c(on),
    label = list(
      on ~ 'Current Antiretroviral medication'
    )
  )
dfipdsnpt %>%
  mutate(
    on = factor(
      on_arv,
      levels = 0:3,
      labels = c('No', 'No, experienced and stopped ART', 'Yes', 'No, restarted ART')
    )
  ) %>%
  tbl_summary(
    include = c(on),
    label = list(
      on ~ 'Current Antiretroviral medication'
    )
  )
dfipdsnpt %>%
  filter(tx_time >= 0) %>% 
  mutate(
    on = factor(
      on_arv,
      levels = 0:3,
      labels = c('No', 'No, experienced and stopped ART', 'Yes', 'No, restarted ART')
    )
  ) %>%
  tbl_summary(
    include = c(on),
    label = list(
      on ~ 'Current Antiretroviral medication'
    )
  )

```
